### 2019-05-16 R/paper 02 Scale-jump analysis of historical retrospective patterns.Rmd
Updated paper; fixed errors in database generator

  # Exclude prediction years (but keep the benchmark information that has no year attached to it)
  filter(is.na(year) | (year <= assessmentyear & year >= 1980) ) %>% 
  filter(assessmentyear >= 1982) %>% 
  filter(is.na(year) | !(is.na(stocksize) & is.na(fishingpressure))) %>%

@@

  # Keep this variable
  filter( variable %in% c("stocksize")) %>%

  arrange(stock, variable, (assessmentyear)) %>% 
  ungroup() %>%

@@

# most recent assessments
d.recent <-
  d.selection %>% 
  filter(assessmentyear == 2018) %>% 
  ungroup() %>% 
  select(year, stock, purpose, variable, value_last=value)

# segments to endpoints
d.segment <-
  d.selection %>% 
  filter( (assessmentyear != 2018 & year == max(year)) ) %>%

@@

# endpoints
d.points <-
  d.selection %>% 
  filter( (assessmentyear != 2018 & year == max(year)) )

@@

    # calculate metrics
    filter(tmp=="firstyear") %>% 
    ungroup() %>% 
    select(stock, assessmentpair, variable, label, value) %>% 
    spread(key=label, value=value) %>% 
    mutate(phi            = (prior-last)/last,
           phi_abs        = abs(phi),
           assessmentyear = as.integer(substr(assessmentpair, 6,9)), 
           scale_jump     = ifelse(phi_abs  > threshold, TRUE, NA),
           scale_jump     = ifelse(phi_abs <= threshold, FALSE, scale_jump)) %>% 
    group_by(stock, variable) %>% 
    mutate(scale_jump     = ifelse(row_number() == 1, NA, scale_jump),
           removelastyears= removelastyears, 
           numberofyears  = numberofyears, 
           threshold      = threshold)

tmp <- pairwise_comparisons(removelastyears = 0, numberofyears=5, threshold = 0.1)

# tmp$d.pcm %>% View()

@@

# plot
d.selection %>%   
  ggplot(aes(x=year, y=value)) +
  theme_publication() +
  theme(
    panel.spacing.x = unit(1, "mm"),
    panel.spacing.y = unit(1, "mm"),
    strip.background = element_blank(),
    strip.text       = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm")),
    legend.position  = "none"
  ) +

  # geom_boxplot(aes(group=year), outlier.shape=NA, colour="gray50", fill=NA) +
  geom_segment(data=d.segment, 
               aes(x=year, xend=year,y=value,yend=value_last), colour="gray", linetype="dashed", size=0.5) +
  geom_line(aes(colour = plot, size=plot, group=assessmentyear)) +
  geom_point(data=d.points) +
  scale_colour_manual(values=c(last = "red", other="black")) +
  scale_size_manual (values=c(last=1, other=0.5)) +
  expand_limits(y=0) +
  scale_y_continuous(breaks = pretty_breaks(),  labels = scales::scientific_format(digits=2)) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(y="ssb (tonnes)") +
  facet_wrap(~stock, scales="free")

@@

res <- pairwise_comparisons(removelastyears = 0, numberofyears=10)

myvar     <- "stocksize"
myspecies <- "mac"
firstyear <- 2008

  filter(variable == myvar) %>% 
  filter(grepl(myspecies, stockkeylabelold)) %>% 
  filter(as.numeric(substr(assessmentpair,1,4)) >= firstyear) %>%

  filter(variable == myvar) %>% 
  filter(grepl(myspecies, stockkeylabelold)) %>% 
  filter(as.numeric(substr(assessmentpair,1,4)) >= firstyear) %>%

  geom_line(data=filter(res$d.pairwise.roughly, 
                        variable == myvar, 
                        grepl(myspecies, stockkeylabelold),
                        as.numeric(substr(assessmentpair,1,4)) >= firstyear), size=0.5) +

  geom_line(data=d.pairwise.means, aes(x=year, y=value, colour=factor(label)), 
            linetype="dashed", lineend="butt") +
  geom_point(data=filter(d.pairwise.means, year==min(year)),
             aes(colour=factor(label)), size=1) +
  geom_point(data=filter(d.pairwise.means, year==max(year)),
             aes(colour=factor(label)), size=1) +

  scale_y_continuous(breaks = pretty_breaks(n=3), labels=scientific_format(digits=1)) +
  scale_x_continuous(breaks = pretty_breaks(n=2)) +
  expand_limits(y=0) +
  labs(x="", y="ssb (tonnes)") +
  facet_wrap( ~ assessmentpair, scales = "free_x", ncol=5)

<!-- ################################################################################## -->
<!-- Plot pairwise comparison means                                                     -->
<!-- ################################################################################## -->

@@

res <- pairwise_comparisons(removelastyears = 0, numberofyears=10, threshold = 0.10)

# threshold <- 0.10

# plot scale jumps between assessments
res$d.pcm %>% 
  distinct(stock, variable, assessmentyear, phi, scale_jump, removelastyears, numberofyears, threshold) %>%
  mutate(    decade = as.character(10*floor(assessmentyear/10)) ) %>% 
  gather(key=metric, value=value, c(phi)) %>%

  ggplot(aes(x=assessmentyear, y=value, group=stock)) +
  theme_publication() +
  theme(
    panel.spacing    = unit(0.1, "lines"),
    panel.grid.major.x = element_line(colour = "gray80"),
    strip.background = element_blank(),
    strip.text.x     = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm")),
    strip.text.y     = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm"), angle = 180),
    strip.placement = "outside",
    legend.position  = "none"
  ) +

  geom_rect(data=filter(d.stockdefinitionchange),
            aes(xmin = y1, xmax=y2, ymin = -1.0, ymax=1.0, fill=rank), 
            alpha=0.4, colour=myPurples[6], inherit.aes = FALSE) +

  geom_dl(data=filter(d.stockdefinitionchange),
          aes(label  = stockkeylabelold, x = y1, y = -0.95),
          method = list(dl.combine("first.points"), rot = 90, cex = 0.6, hjust=0.0, vjust=0.9),
          inherit.aes = FALSE, colour=brewer.pal(6, "Purples")[6]) +

  scale_fill_manual(values = brewer.pal(6, "Purples")[-(1:2)]) +

  labs(x=" ", y="") +
  facet_wrap(~stock)

res <- pairwise_comparisons(removelastyears = 0, numberofyears=10)

threshold <- 0.10

# plot scale jumps between assessments vs assessment methods

res$d.pcm %>% 
  distinct(stock, variable, assessmentyear, phi, scale_jump, removelastyears, numberofyears, threshold) %>%
  mutate(    decade = as.character(10*floor(assessmentyear/10)) ) %>% 
  gather(key=metric, value=value, c(phi)) %>%

  ggplot(aes(x=assessmentyear, y=value, group=stock)) +
  theme_publication() +
  theme(
    panel.spacing    = unit(0.1, "lines"),
    panel.grid.major.x = element_line(colour = "gray80"),
    strip.background = element_blank(),
    strip.text.x     = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm")),
    strip.text.y     = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm"), angle = 180),
    strip.placement = "outside",
    legend.position  = "none"
  ) +

  geom_rect(data=filter(d.assessmentmodels),
            aes(xmin = y1, xmax=y2, ymin = -1.0, ymax=1.0, fill=rank), 
            alpha=0.4, colour=myBlues[6], inherit.aes = FALSE) +

  geom_line (aes(colour=factor(decade))) +
  geom_point(aes(colour=factor(decade), size = ifelse( abs(value) > threshold, 0.3, 0.0001))) +

  labs(x=" ", y="") +
  facet_wrap(~stock)

res <- pairwise_comparisons(removelastyears = 0, numberofyears=10, threshold=0.1)

# plot scale jumps between assessments vs assessment methods

res$d.pcm %>% 
  distinct(stock, variable, assessmentyear, phi, scale_jump, removelastyears, numberofyears, threshold) %>%
  mutate(    decade = as.character(10*floor(assessmentyear/10)) ) %>% 
  gather(key=metric, value=value, c(phi)) %>%

  ggplot(aes(x=assessmentyear, y=value, group=stock)) +
  theme_publication() +
  theme(
    panel.spacing    = unit(0.1, "lines"),
    panel.grid.major.x = element_line(colour = "gray80"),
    strip.background = element_blank(),
    strip.text.x     = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm")),
    strip.text.y     = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm"), angle = 180),
    strip.placement = "outside",
    legend.position  = "none"
  ) +

  geom_rect(data=filter(d.stockdefinitionchange),
            aes(xmin = y1, xmax=y2, ymin = 0.00, ymax=1.0), 
            alpha=0.4, fill=myPurples[6], colour=myPurples[6], inherit.aes = FALSE) +

  geom_rect(data=filter(d.assessmentmodels),
            aes(xmin = y1, xmax=y2, ymin = -1.0, ymax=0.0, fill=rank), 
            alpha=0.4, colour=myBlues[6], inherit.aes = FALSE) +

  labs(x=" ", y="") +
  facet_wrap(~stock)

@@

```{r jumpratio, echo=FALSE, fig.asp=0.5, fig.align="center", message=FALSE, warning=FALSE}

```{r jumpratio, echo=FALSE, fig.asp=0.5, fig.align="center", message=FALSE, warning=FALSE}

>>>>

  # Exclude prediction years (but keep the benchmark information that has no year attached to it)
  filter(is.na(year) | (year <= assessmentyear) ) %>% 
  filter(assessmentyear >= 1982) %>% 
  filter(assessmentyear <= 2018) %>% 
  filter(is.na(year) | !(is.na(stocksize) & is.na(fishingpressure))) %>%

@@

  # Keep this variable
  filter( variable %in% c("stocksize", "recruitment", "fishingpressure")) %>%

  arrange(stock, assessmentyear, variable, year) %>% 
  ungroup() %>%

@@

# most recent assessments
d.recent <-
  d.selection %>% 
  ungroup() %>% 
  filter(assessmentyear == max(assessmentyear, na.rm=TRUE)) %>% 
  ungroup() %>% 
  select(year, stock, purpose, variable, value_last=value)

# segments to endpoints
d.segment <-
  d.selection %>% 
  ungroup() %>% 
  filter( (assessmentyear != max(assessmentyear, na.rm=TRUE) ) ) %>%
  group_by(stock, assessmentyear, purpose, variable) %>% 
  filter(year == max(year)) %>%

@@

# endpoints
d.points <-
  d.segment %>% 
  filter( (assessmentyear != 2018 & year == max(year)) )

# removelastyears=0; numberofyears=c(5,10); threshold=0.10

  # rly <- 0
  # ny <- 5
  # th <- 0.20

  d.pairwise.roughly <- data.frame()

  for (rly in removelastyears) {
    for (ny in numberofyears) {

      print(paste("removelastyears:",rly, "numberofyears:",ny,sep=" "))

      # pairwise data roughly
      d.pairwise.roughly <-
        bind_rows(d.pairs.prior, d.pairs.last) %>%

        # filter the appropriate period
        group_by(stock, assessmentpair, variable) %>% 
        # filter(year <= as.numeric(substr(assessmentpair,1,4)) - rly) %>% 
        filter(year >= as.numeric(substr(assessmentpair,1,4)) - rly - ny -5) %>% 
        arrange(stock, assessmentpair, assessmentyear, variable) %>% 
        mutate(removelastyears = rly, 
               numberofyears   = ny) %>%

        # add actual yearrange used
        group_by(stock, assessmentpair, assessmentyear, variable, removelastyears, numberofyears) %>%
        mutate(yearrange      = paste(as.character(min(year)),as.character(max(year)), sep="-"),
               nyears         = n_distinct(year)) %>%

        # add to dataframe
        bind_rows(d.pairwise.roughly)

    } # end of numberofyears
  } # end of removelastyears

  # pairwise data
  d.pairwise <-
    d.pairwise.roughly %>%
    group_by(stock, assessmentpair, variable, removelastyears, numberofyears) %>%

    # filter the appropriate period
    filter(
      year <= as.numeric(substr(assessmentpair,1,4)) - removelastyears,
      year >= as.numeric(substr(assessmentpair,1,4)) - removelastyears - (numberofyears-1)) %>%

    # sort
    arrange(stock, assessmentpair, assessmentyear, removelastyears, numberofyears, year) %>%

    # filter only years that are in common in last and prior
    group_by(stock, assessmentpair, variable, removelastyears, numberofyears, year) %>%
    mutate(n = n()) %>%
    filter(n() == 2) %>%

    # add actual yearrange used
    group_by(stock, assessmentpair, variable, removelastyears, numberofyears) %>%
    mutate(yearrange      = paste(as.character(min(year)),as.character(max(year)), sep="-"),
           nyears         = n_distinct(year))

  # start with empty data.frame
  d.pcm <- data.frame()
  # th <- 0.2

  # loop over thresholds
  for (th in threshold) {

    print(paste("threshold:",th,sep=" "))

      # calculate means per period
      group_by(stock, assessmentpair, variable, removelastyears, numberofyears, yearrange, nyears) %>%
      mutate(   firstyear      = min(year),
                lastyear       = max(year)) %>%
      group_by(stock, assessmentpair, variable, assessmentyear, label, yearrange, nyears,
               removelastyears, numberofyears, firstyear, lastyear) %>%
      summarise(value = mean(value, na.rm=TRUE)) %>%
      gather(key=tmp, value=year, firstyear:lastyear) %>%
      arrange(stock, assessmentpair, variable, assessmentyear, yearrange, label) %>%

      # calculate metrics
      filter(tmp=="firstyear") %>%
      ungroup() %>%
      dplyr::select(-tmp, -assessmentyear) %>%
      spread(key=label, value=value) %>%
      mutate(phi            = (prior-last)/last,
             phi_abs        = abs(phi),
             assessmentyear = as.integer(substr(assessmentpair, 6,9)),
             scale_jump     = ifelse(phi_abs  > th, TRUE, NA),
             scale_jump     = ifelse(phi_abs <= th, FALSE, scale_jump),
             threshold      = th) %>%
      group_by(stock, variable, removelastyears, numberofyears, threshold) %>%
      mutate(scale_jump     = ifelse(row_number() == 1, NA, scale_jump)) %>%

      bind_rows(d.pcm)

  } # end of threshold

res <- pairwise_comparisons(removelastyears = c(0,1), numberofyears=c(2, 5,10), threshold = c(0.1, 0.2))

# res$d.pairwise.roughly %>% 
#   filter(grepl("mac", stock), variable=="stocksize", numberofyears==5, removelastyears==1,
#          assessmentpair=="2017-2018") %>% View()

# res$d.pairwise %>% 
#   filter(grepl("mac", stock), variable=="stocksize", numberofyears==10, removelastyears==1,
#          assessmentpair=="2017-2018") %>% View()

# res$d.pcm %>% View()
# res$d.pairwise.roughly %>% View()
# res$d.pairwise %>% View()

# d.pairwise %>% 
#   filter(grepl("mac", stock), assessmentpair == "2017-2018") %>% 
#   filter(removelastyears==myremoveyears, numberofyears==mynyears) %>%  
#   filter(variable == myvar) %>% 
#   View()

# res$d.pcm %>% ungroup() %>% distinct(threshold)

myvar <- "stocksize"

# plot
d.selection %>%  
  filter(variable==myvar) %>% 
  ggplot(aes(x=year, y=value)) +
  theme_publication() +
  theme(
    panel.spacing.x = unit(1, "mm"),
    panel.spacing.y = unit(1, "mm"),
    strip.background = element_blank(),
    strip.text       = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm")),
    legend.position  = "none"
  ) +

  # geom_boxplot(aes(group=year), outlier.shape=NA, colour="gray50", fill=NA) +
  geom_segment(data=filter(d.segment, variable==myvar), 
               aes(x=year, xend=year,y=value,yend=value_last), colour="gray", linetype="dashed", size=0.5) +
  geom_line(aes(colour = plot, size=plot, group=assessmentyear)) +
  # geom_point(data=d.points) +
  geom_point(data=filter(d.segment, variable==myvar), 
               aes(x=year, y=value), colour="black", size=1) +

  scale_colour_manual(values=c(last = "red", other="black")) +
  scale_size_manual (values=c(last=1, other=0.5)) +
  expand_limits(y=0) +
  scale_y_continuous(breaks = pretty_breaks(),  labels = scales::scientific_format(digits=2)) +
  scale_x_continuous(breaks = pretty_breaks(n=3)) +
  labs(y=myvar) +
  facet_wrap(~stock, scales="free")

myvar         <- "stocksize"
myspecies     <- "her"
firstyear     <- 2008
firstplotyear <- 2000
mynyears      <- 5
myremoveyears <- 0
mythreshold   <- 0.1
mytitle       <- paste0(toupper(myspecies), ", ", myvar, ", number of years: ",mynyears, 
                        ", removed years: ", myremoveyears)

  res$d.pairwise %>% 
  # ungroup() %>% 
  # distinct(removelastyears, numberofyears, threshold) %>% 
  # View()

  filter(removelastyears==myremoveyears, numberofyears==mynyears) %>% 
  filter(variable == myvar) %>% 
  filter(grepl(myspecies, stockkeylabelold)) %>% 
  filter(as.numeric(substr(assessmentpair,1,4)) >= firstyear) %>%

  filter(removelastyears==myremoveyears, numberofyears==mynyears) %>% 
  filter(variable == myvar) %>% 
  filter(grepl(myspecies, stockkeylabelold)) %>% 
  filter(as.numeric(substr(assessmentpair,1,4)) >= firstyear) %>%

  geom_line(data=filter(res$d.pairwise.roughly, 
                        variable == myvar, 
                        grepl(myspecies, stockkeylabelold),
                        as.numeric(substr(assessmentpair,1,4)) >= firstyear,
                        year >= firstplotyear), 
            size=0.5) +

  geom_line(data=d.pairwise.means, aes(x=year, y=value, colour=factor(label)),
            linetype="dashed", lineend="butt", inherit.aes = FALSE) +
  geom_point(data=filter(d.pairwise.means, year==min(year)),
             aes(colour=factor(label)), size=1) +
  geom_point(data=filter(d.pairwise.means, year==max(year)),
             aes(colour=factor(label)), size=1) +
  geom_text (data=filter(res$d.pcm,
                         variable == myvar,
                         grepl(myspecies, stock),
                         as.numeric(substr(assessmentpair,1,4)) >= firstyear,
                         numberofyears == mynyears,
                         removelastyears == myremoveyears,
                         threshold == mythreshold),
             aes(x=firstplotyear, y=0, label = paste0("phi=", round(phi_abs, digits=2))), 
             hjust=0, vjust=0, nudge_x = -0.1, nudge_y=0.1,  inherit.aes = FALSE) +

  scale_y_continuous(breaks = pretty_breaks(n=3), labels=scientific_format(digits=1)) +
  scale_x_continuous(breaks = pretty_breaks(n=2)) +
  expand_limits(y=0) +
  labs(x="", y=myvar, title=mytitle) +
  facet_wrap( ~ assessmentpair, ncol=5)

# names(res)
# filter(res$d.pcm,
#                          variable = myvar,
#                          grepl(myspecies, stock),
#                          as.numeric(substr(assessmentpair,1,4)) >= firstyear,
#                          numberofyears == mynyears,
#                          removelastyears == myremoveyears,
#                          threshold == mythreshold)
```

<!-- ################################################################################## -->
<!-- Plot scale jumps                                                     -->
<!-- ################################################################################## -->

```{r echo=FALSE, fig.align="center", fig.asp=1.0, message=FALSE, warning=FALSE}

# res <- pairwise_comparisons(removelastyears = 0, numberofyears=10, threshold = 0.10)

# threshold <- 0.10
mynyears      <- 10
myremoveyears <- 0
mythreshold   <- 0.1
myvar         <- "stocksize"
mytitle       <- paste0(myvar, ", number of years: ",mynyears, ", threshold: ", 
                        comma_format(digits=1, suffix="%")(100*mythreshold))

# res$d.pcm %>% View()
# res$d.pcm %>% ungroup() %>% distinct(threshold)

# plot scale jumps between assessments
t <-
  res$d.pcm %>% 
  filter(variable        == myvar,
         numberofyears   == mynyears,
         removelastyears == myremoveyears,
         threshold       == mythreshold) %>% 
  distinct(stock, variable, assessmentyear, phi, scale_jump, removelastyears, numberofyears, threshold) %>%
  mutate(    decade = as.character(10*floor(assessmentyear/10)) ) %>% 
  gather(key=metric, value=value, c(phi))

t %>%   
  ggplot(aes(x=assessmentyear, y=value, group=stock)) +
  theme_publication() +
  theme(
    panel.spacing    = unit(0.1, "lines"),
    panel.grid.major.x = element_line(colour = "gray80"),
    strip.background = element_blank(),
    strip.text.x     = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm")),
    strip.text.y     = element_text(face="bold", hjust=0, margin = margin(2,0,2,0, "mm"), angle = 180),
    strip.placement = "outside",
    legend.position  = "none"
  ) +

  # stock definition change
  geom_rect(data=filter(d.stockdefinitionchange),
            aes(xmin = y1, xmax=y2, ymin = 0.00, ymax=1.0), 
            alpha=0.4, fill=myPurples[6], colour=myPurples[6], inherit.aes = FALSE) +

  # assessmentmodel change
  geom_rect(data=filter(d.assessmentmodels),
            aes(xmin = y1, xmax=y2, ymin = -1.0, ymax=0.0, fill=rank), 
            alpha=0.4, colour=myBlues[6], inherit.aes = FALSE) +

  labs(x=" ", y="", title=mytitle) +
  facet_wrap(~stock)

```{r, echo=FALSE, fig.asp=0.8, fig.align="center", message=FALSE, warning=FALSE}

myremoveyears <- 0
myvar         <- "stocksize"
myspecies     <- "cod"
mytitle       <- paste0(toupper(myspecies), ", ", myvar, ", removed years: ", myremoveyears)

res$d.pcm %>%

  filter(grepl(myspecies, stock)) %>% 
  filter(variable        == myvar) %>% 
  filter(removelastyears == myremoveyears) %>%

  distinct(stock, variable, assessmentyear, phi, scale_jump, removelastyears, numberofyears, threshold) %>%
  mutate(    decade = as.character(10*floor(assessmentyear/10)) ) %>% 
  gather(key=metric, value=value, c(phi)) %>%

  labs(x=" ", y="", title=mytitle) +
  facet_grid(threshold ~ numberofyears)

```{r jumpratio, eval=FALSE, fig.align="center", fig.asp=0.5, message=FALSE, warning=FALSE, include=FALSE}

# NOT SHOWN, NOT USED

```{r jumpratio, eval=FALSE, fig.align="center", fig.asp=0.5, message=FALSE, warning=FALSE, include=FALSE}

# NOT SHOWN, NOT USED

```{r eval=FALSE, include=FALSE}

# NOT SHOWN, NOT USED

@@

@@

@@

@@

# NOT SHOWN

@@
